---
title: "Assignment #5 - Graphing Realtionships"
author: "Matteo Larrode"
format: 
  gfm:
    code-fold: true
    prefer-html: true
---

## Question 1: 2016 and 2020 Democratic vote shares

Let's first load the packages and data we will need for this assignment

```{r}
#| output: false

library(tidyverse)
library(haven)
library(kableExtra) #table
library(psych) #describe function
library(ggplot2)
library(ggthemes)
library(ggrepel) #labeling state names
library(scales)

vote_2020 <- read_csv("data/us_vote_2020.csv")
dem_share_80_16 <- read_dta("data/leipvote1980_2016wide.dta")

```

Now I join the 2020 election data to democratic votes shares of the two party vote, from 1980 to 2016 to the 1980-2016.

```{r}

vote_80_20 <- left_join(dem_share_80_16, vote_2020, by = "state") %>%
  mutate(pctdem2020 = dem_percent / 100)

vote16_20 <- vote_80_20 %>%
  select(state, stateid, pctdem2016, pctdem2020)

```

Here is a table of summary statistics for the 2016 and 2020 Democratic vote shares.

```{r}
summary_stats <- describe(vote16_20[ , c('pctdem2016', 'pctdem2020')], fast = TRUE) %>%
  as.data.frame() %>%
  select(-vars) %>%
  mutate(across(-n, ~ . * 100))
  
rownames(summary_stats) <- c("2016", "2020")

# Create HTML table
table1 <- kbl(summary_stats, format = "html", align = "c", caption = "Summary Statistics for Democratic Vote Shares (2016-2020)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = F,
                fixed_thead = T)

table1
```

We can see that there was a slight increase, on average, in the percentage of the two-party vote that went to the Democratic presidential candidate from 2016 to 2020. The mean percentage increased from 47.99% in 2016 to 48.65% in 2020, and the minimum percentage increased too. The standard deviation and range decreased as well, suggesting that there was slightly less variability in the percentage of the two-party vote that went to the Democratic presidential candidate across all states in 2020 compared to 2016.

## Question 2: Scatter plot

Now let us generate the vote swing scatter plot with the 2020 Democratic percentage of the vote on the Y axis and the 2016 Democratic percentage of the vote on the X axis. 

```{r, dpi = 150}
#| warning: false

plot1 <- ggplot(vote16_20, aes(x = pctdem2016*100, y = pctdem2020*100)) +
  geom_point() +
  geom_text_repel(aes(label = state),
                  max.overlaps = Inf) +
  theme(text=element_text(family="Roboto"),
        panel.background = element_blank(),
        plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"),
        plot.title=element_text(hjust=0.5, face="bold", size=12, margin = margin(b=10)),
        plot.caption = element_text(size=9, vjust = -2),
        axis.ticks = element_blank(),
        panel.grid.major = element_line(color = "gray", linetype = "dashed", linewidth = 0.2))+
  labs(x = "2016 Democratic Vote %", 
       y = "2020 Democratic Vote %",
       title = "Vote for the Democratic Presidential Candidate in 2016 and 2020",
       caption = "@matteoStats") + # axis labels
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") # add 45 degree line


plot1
```
Even using the {ggrepel} package to label state names, the plot is too crowded! In order to be able to read and describe the graph, let us try labeling only 10 states. To bring more insight on states that experienced change, I will choose those with the highest difference between 2016 and 2020. 

```{r}

vote16_20_new <- vote16_20 %>%
  mutate(pct_diff = abs(pctdem2020 - pctdem2016),
         #use rank function
         top_10 = ifelse(rank(desc(pct_diff)) <= 10, TRUE, FALSE))

```

Now let us plot the new graph, with less state labels.

```{r, dpi = 150}
#| warning: false

plot2 <- ggplot(vote16_20_new, aes(x = pctdem2016*100, y = pctdem2020*100, col = top_10)) +
  geom_point() +
  geom_text_repel(data = filter(vote16_20_new, top_10),
                  aes(label = state),
                  #get labels away from the points
                  position = position_nudge_repel(y = 1)) +
  theme(text=element_text(family="Roboto"),
        panel.background = element_blank(),
        plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"),
        plot.title=element_text(hjust=0.5, face="bold", size=12, margin = margin(b=10)),
        plot.caption = element_text(size=9, vjust = -2),
        axis.ticks = element_blank(),
        panel.grid.major = element_line(color = "gray", linetype = "dashed", linewidth = 0.2),
        legend.position = "none")+
  scale_color_manual(values = c("grey", "black"))+
  labs(x = "2016 Democratic Vote %", 
       y = "2020 Democratic Vote %",
       title = "Vote for the Democratic Presidential Candidate in 2016 and 2020",
       caption = "@matteoStats") + # axis labels
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") # add 45 degree line


plot2
```
The points above the 45-degree line indicate a gain in Democratic votes between 2016 and 2020, while points below the line indicate a loss in Democratic votes. From the plot, we can see that most states had a higher percentage of Democratic votes in 2020 compared to 2016, as indicated by the majority of points above the 45-degree line. Some notable exceptions include California, and Hawaii, where the Democratic vote share decreased from 2016 to 2020. 

The choice of states to label does leave out some information. For example, it is interesting to note that some traditionally Republican states, such as Arizona and Georgia, had a significant gain in Democratic votes in 2020. Overall, this plot provides a useful visual representation of the changing political landscape in the United States between the 2016 and 2020 presidential elections.

At the top-right corner of the plot, the state of Washington, D.C. seems like an outlier. For both 2016 and 2020, it has a much higher percentage of vote for the Democrat candidate than the other states. Let us try to omit this state for the next plot. 

```{r, dpi = 150}
#| warning: false

plot3 <- ggplot(data = filter(vote16_20_new, state != "District of Columbia"),
                aes(x = pctdem2016*100, y = pctdem2020*100, col = top_10)) +
  geom_point() +
  geom_text_repel(data = filter(vote16_20_new, top_10 & state != "District of Columbia"),
                  aes(label = state),
                  #get labels away from the points
                  position = position_nudge_repel(y = 1)) +
  theme(text=element_text(family="Roboto"),
        panel.background = element_blank(),
        plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"),
        plot.title=element_text(hjust=0.5, face="bold", size=12, margin = margin(b=10)),
        plot.caption = element_text(size=9, vjust = -2),
        axis.ticks = element_blank(),
        panel.grid.major = element_line(color = "gray", linetype = "dashed", linewidth = 0.2),
        legend.position = "none")+
  scale_color_manual(values = c("grey", "black"))+
  labs(x = "2016 Democratic Vote %", 
       y = "2020 Democratic Vote %",
       title = "Vote for the Democratic Presidential Candidate in 2016 and 2020 (without DC)",
       caption = "@matteoStats") + # axis labels
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") # add 45 degree line


plot3
```

The effect of removing Washington D.C. is that the points of the scatter plot are less tightly clustered. Removing the outlier - DC - helps to get a clearer view of the evolution of the vote for the Democrat candidate between 2016 and 2020 in the other states. 


## Question 3

From our data on democratic vote from 1980 to 2020, we filter out Washington D.C., and reshape the data from wide form to long form. 

```{r}
vote_80_20_new <- vote_80_20 %>%
  filter(state != "District of Columbia") %>%
  pivot_longer(cols = starts_with("pctdem"), 
               names_to = "Year", 
               values_to = "vote_pct")%>%
  select(state, stateid, Year, vote_pct)%>%
  mutate(Year = str_extract(Year, "\\d{4}")) %>%
  print(n = 15)
```
Now, we can generate a line plot for the 50 states from 1980 to 2020, graphing percent Democratic on the y-axis and year on the x-axis. To bring more insight, we will emphasize the 5 states with the highest variation, using standard deviation.

A bit of setup is necessary first. 

```{r}
#selection of the colors for highlighted lines
highlight_colors <- c("#17648d", "#51bec7","#008c8f","#d6ab63", "#843844")

#find the highest and lowest averages
state_sd <- vote_80_20_new %>%
  group_by(state)%>%
  summarize(sd = sd(vote_pct))

#selection of 6 states
names(highlight_colors) <- c('West Virginia', 
                             'Vermont', 
                             'New Hampshire',
                             'Hawaii',
                             'Arkansas')


#new data frame with only those 5 focus areas
highlight_data <- vote_80_20_new %>%
  filter(state %in% names(highlight_colors))
```

Now on to the plotting! 

```{r, dpi = 150}

plot4 <- ggplot(vote_80_20_new, aes(x = Year, y = vote_pct, group = state)) +
  geom_line(linewidth = 0.75, color = '#d4dddd') +
  geom_line(data = highlight_data, aes(color = state), linewidth = 1.25) + 
  
  labs(title = "Democratic vote share by state for every Presidential election (1980 - 2020), \n excluding Washington D.C.",
       caption = "@matteoStats",
       x = NULL,
       y = NULL)+
  
  scale_y_continuous(position = "right",
                     labels = scales::percent_format())+
  scale_colour_manual(values = highlight_colors) + 
  
  theme(aspect.ratio = 3.2/6,
        text = element_text(family = "Roboto Condensed"),
        plot.margin = margin(t = 0, r = 0.75, b = 0, l = 0.75, unit = "cm"),
        plot.background = element_rect(fill = "white"),
        panel.background = element_rect(fill = "white"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "#dcdbd8"),
        panel.grid.minor.y = element_blank(),
        legend.text = element_text(margin = margin(l=3), size = 10),
        legend.title = element_blank(),
        legend.position = c(0.15, 0.9),
        legend.key.width = unit(25,"pt"),
        legend.key.height = unit(15, "pt"),
        legend.key = element_blank(),
        axis.text = element_text(size = 9, color = "gray8"),
        axis.line.x  = element_line(color = "gray8"),
        axis.ticks.y = element_blank(),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
        plot.caption = element_text(hjust = 1, size = 11, colour = "#4B4B4B"))

plot4
```
Many lines in the plot seem to generally show a gradual increase in the percentage of Democratic voters over time; however, this does not necessarily mean that voting is becoming more Democratic. In fact, the data could suggest a growing polarization between Democratic and Republican voters. This is particularly evident when observing the widening of the spread of Democratic vote share over the years, at the national level.

There are some notable spikes and dips in the lines, indicating abrupt changes in voting patterns. For example, many states experienced an important increase in Democratic voters in the 1990s, and a subsequent decrease at the turn of the millennium.

The lines for each state are generally quite flat, indicating that the percentage of Democratic voters does not vary greatly within states over time, with some states that consistently vote more, or less, Democratic others.

This leads me to the story told by my choice to highlight the states with the highest standard deviation in democratic voting share. A state with a high standard deviation indicates that it experiences more variability in the percentage of Democratic voters across time, compared to states with lower standard deviations. Those highlighted states go through noteworthy changes in demographics or political attitudes over time, leading to more variability in the percentage of Democratic voters. Identifying those states is especially important in the United States, due to its voting system. States with high standard deviation are potentially more competitive in elections, with voters more likely to switch between Democratic and Republican candidates, having major implications for election strategies and the allocation of campaign resources by political parties.


Now let's try faceting the graph by census region. 

```{r, dpi = 150}
#install.packages("usa")
library(usa)

usa_region <- usa::states %>%
  select(name, region)

#attach census region to dataset
vote_80_20_new <- left_join(vote_80_20_new, usa_region, by = c("state" = "name"))

highlight_data <- vote_80_20_new %>%
  filter(state %in% names(highlight_colors))

#plot
plot5 <- ggplot(vote_80_20_new, aes(x = Year, y = vote_pct, group = state)) +
  geom_line(linewidth = 0.5, color = '#d4dddd') +
  geom_line(data = highlight_data, aes(color = state), linewidth = 1) + 
  
  labs(title = "Democratic vote share by state for every Presidential election (1980 - 2020), \n excluding Washington D.C.",
       caption = "@matteoStats",
       x = NULL,
       y = NULL)+
  
  scale_y_continuous(position = "right",
                     labels = scales::percent_format())+
  scale_colour_manual(values = highlight_colors) + 
  
  facet_wrap(~ region, ncol=2) +

  
  theme(aspect.ratio = 3.2/6,
        text = element_text(family = "Roboto Condensed"),
        plot.margin = margin(t = 0, r = 0.5, b = 0, l = 0.5, unit = "cm"),
        plot.background = element_rect(fill = "white"),
        panel.background = element_rect(fill = "white"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "#dcdbd8"),
        panel.grid.minor.y = element_blank(),
        legend.text = element_text(margin = margin(l=3), size = 10),
        legend.title = element_blank(),
        legend.position = "bottom",
        legend.key.width = unit(25,"pt"),
        legend.key.height = unit(15, "pt"),
        legend.key = element_blank(),
        axis.text = element_text(size = 7, color = "gray8"),
        axis.line.x  = element_line(color = "gray8"),
        axis.ticks.y = element_blank(),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
        plot.caption = element_text(hjust = 1, size = 11, colour = "#4B4B4B"))

plot5
```
Overall, this faceted graph suggests that regional differences play an important role in shaping the political landscape of the United States. The Northeast and West regions tend to vote more for Democrat presidential candidates than the Midwest and South regions.  Additionally, the chart indicates that the percentage of Democratic voters has been relatively stable in some regions while it has fluctuated more in others (e.g., West).

## Question 4: Final Project Data

I am passionate about both political science and technology, so I grew interested in the effect of the rise and diffusion of new technologies on international relations. One particular topic that fascinates me is the relationship between internet use, especially activity on social networks, and freedom of speech. 

Social media platforms have fundamentally transformed the landscape of social and political movements. They have allowed individuals to bypass traditional gatekeepers of information, such as the mainstream media, and connect with each other directly. This has enabled dissidents to spread information and organize more quickly and effectively than ever before. But on the other hand, they can also be easily monitored and censored by governments, potentially putting activists at risk.

I created a dataset called "final_data.csv" by joining data on Internet penetration, provided by the [International Telecommunication Union (ITU)](https://www.itu.int/en/ITU-D/Statistics/Pages/stat/default.aspx), and around 50 different metrics used by [V-Dem](https://www.v-dem.net/data/the-v-dem-dataset/) to measure levels of democracy. 

```{r}
df <- read_csv("data/final_data.csv")%>% 
  drop_na()

head(df)
```

My main research question is: "Does higher Internet penetration lead to high levels of freedom of speech?"

More specifically, could the Internet empower people who are usually less able to express themselves to finally share their opinions? Here is a scatterplot that illustrates the relationship between freedom of discussion for women, my outcome variable, and access to the Internet, my main independent variable?


```{r, dpi = 150}

free_disc_women_plot <- ggplot(df, aes(x = value_2021_perc, y = v2cldiscw))+
  geom_point()+
  geom_smooth(method = "lm")+

  theme(text=element_text(family="Roboto"),
        panel.background = element_blank(),
        plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm"),
        plot.title=element_text(hjust=0.5, face="bold", size=12, margin = margin(b=10)),
        plot.caption = element_text(size=9, vjust = -2),
        axis.text.x = element_text(vjust = 5),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid.major = element_line(color = "gray", linetype = "dashed", linewidth = 0.2))+
  scale_x_continuous(name ="Internet Penetration",labels = scales::percent_format(accuracy = 1))+
  scale_y_continuous(name = "Freedom of Discussion for Women")+
  labs(title = "Higher Access to the Internet is Correlated to More Freedom of Discussion for Women",
       caption="Source: ITU, V-Dem | @matteoStats")

free_disc_women_plot

```
From this simple scatterplot, and the linear regression fitted to the data, it seems that countries with high levels of Internet penetration are associated to high levels of freedom of discussion for women. However, an important omitted variable bias could emerge if we do not take governance into account. Indeed, democracies, usually characterized by high levels of freedom of discussion, are also often more developed than authoritarian regimes, and therefore have higher Internet penetration rates on average. 

An interesting extra step to analyze the relationship between Internet penetration and freedom of discussion could be to control for the level of democracy. To reach this objective, we will use V-Dem's Liberal democracy index (v2x_libdem), that measures to what extent is the ideal of liberal democracy achieved. To clarify, the liberal principle of democracy emphasizes the importance of protecting individual and minority rights against the tyranny of the state and the tyranny of the majority. This is achieved by constitutionally protected civil liberties, strong rule of law, an independent judiciary, and effective checks and balances that, together, limit the exercise of executive power. 

To explore this addition of an independent variabe to the model, we can created the same scatterplot as before, but faceted by different levels of the liberal democracy index. 


