---
title: "Assignment #5 - Graphing Realtionships"
author: "Matteo Larrode"
format: 
  gfm:
    code-fold: true
editor: visual
---

## Question 1: 2016 and 2020 Democratic vote shares

Let's first load the packages we will need for this question

```{r}
#| output: false
library(tidyverse)

```

1) Left join the 2020 election data we used in challenge 7_1 (“us_vote_2020.csv”) to the 1980-2016 data frame. It is also posted on the assignment page. Create a dempct2020 variable using the dem_votes and rep_votes variables and a mutate statement. [Political science aside: what we call the “two party vote” limits the analysis to votes for Democrats and Republicans only.]

```{r}

```

2) Create a table of summary statistics for the 2016 and 2020 Democratic vote shares. Describe the overall pattern, in one sentence.


## Question 2

Using the data from 1, generate a vote swing scatter plot with the 2020 Democratic percentage of the vote on the Y axis and the 2016 Democratic percentage of the vote on the X axis, labeling the points with the state names, and adding a 45 degree line to the plot using geom_abline().
Describe your graph. What is is overall pattern in the data?
What challenges are created by labeling the states? Try labeling only some of the states, justifying your choice of states.
Are there unusual states in the plot? If so, which ones?
Make a graph that omits Washington D.C. from the data. What effect does that have on the structure of the graph?


```{r}

```


## Question 3

Again using the data from 1, filter out Washington D.C., reshape the data from wide form to long form using pivot_longer() and generate a line plot for the 50 states from 1980 to 2020, graphing percent Democratic on the y-axis and year on the x-axis. Choose a set of states that you find interesting, and emphasize those states using colors.
Describe the overall patterns in the data over time. What central features are particularly noteworthy, and why?
What story are you telling with your use of color to highlight a set of states?
Remember that in section 7.1, Healy faceted his election graph by the four census regions. Generate a second graph structured as before, this time faceting by census region.

```{r}


```


## Question 4: Final Project Data

Now using some data that you would like to use in your final project, answer the following:
Describe in detail the source of your data.
What question do you want to ask? That is, what is your research question, or, put differently, what question will your graph answer?
Create a scatterplot (or alternative graph, as appropriate) that illustrates the relationship between your outcome of interest and your main independent variable of interest. Describe the graph in a paragraph. What story does the graph convey?


```{r, dpi = 150}


```
